---
title: "HW 7: Hypothesis Testing"
subtitle: "Math 141, Week 8"
author: "Insert Your Name Here"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---


## Assigned: Friday, March 20th
## Due: Friday, April 3rd at 11:59pm on Gradescope

In this homework assignment you will:

- Demonstrate proficiency with the concepts: null and alternative hypothesis, p-value, type I and II errors, power, and how these relate to specific research questions
- Consider some pitfalls associated with p-values
- Derive confidence intervals from bootstrap distributions with both percentile and standard error methods
- Practice creating and interpreting confidence intervals with varying levels of confidence
- Interpret the probabilistic meaning of confidence intervals and avoid misconceptions


### Add any collaborators here!

I collaborated with...

**Remember that you are not allowed to use any generative AI models (like ChatGPT) on your assignment**



## Exercise 1: Percentile-Method Confidence Intervals

Suppose you calculate a bootstrap distribution based on 10,000 re-samples from a given sample. The bootstrap distribution is shown below, as well as the 0.025, 0.05, 0.25, 0.75, 0.95, and 0.975 quantiles of the bootstrap distribution (shown as vertical lines in the histogram as well).

```{r fig.asp = 0.3, echo = F}
bs_stats <- data.frame(x = rnorm(10000, 10 / 50, 0.05656854))
quantiles <- round(quantile(bs_stats$x,
                            c(0.025, 0.05, 0.25, 0.75, 0.95, 0.975)), 3)
ggplot(bs_stats, aes(x)) +
  geom_histogram(color = "white", bins = 25) +
  scale_x_continuous(limits = c(0, .4)) + theme_bw() +
  labs(title = "Bootstrap distribution based on sample of size n = 50",
       x = "Bootstrap sample statistics") +
  geom_vline(xintercept = quantiles,
             color = "red",
             linetype = 2)
quantiles
```



**a.** Based on the histogram and table of quantiles, write down an appropriate confidence interval based on a (i) 95% confidence level, (ii) 90% confidence interval, and (iii) 50% confidence interval.

********************************





********************************



**b.** What happens to the confidence interval width as the confidence level gets higher? Why?

********************************






********************************




**c.** Would a 100% confidence interval be useful? 

********************************






********************************


**d.** Suppose the norm in our discipline is to use 95% confidence intervals, and that our research collaborator is unhappy with how wide our confidence interval from (a) is. **Would increasing the number of bootstrap replicates help? Would taking more samples to increase our sample size help?**

********************************






********************************



\newpage




## Exercise 2: Confidence Interval Misconceptions I 

You are trying to find out the true proportion of Oregon residents who got COVID between July 15 and October 15. Based on my bootstrap distribution, suppose I told you that an appropriate 99% confidence interval was (0.053, 0.349). Consider the following two interpretations:

- Interpretation A: There is a 99% chance that the true proportion of Oregon residents who got COVID between July 15 and October 15 is between 0.053 and 0.349.

- Interpretation B: For 99% of samples we could have collected, the true parameter will be inside the corresponding 99% confidence interval; we are 99% confident that the true proportion of Oregon residents who got COVID between July 15 and October 15 is between 0.053 and 0.349.

Only one of these is correct.



**a.** How do these interpretations differ? 

********************************






********************************



**b.** Which interpretation is correct, and why?

********************************







********************************




\newpage





## Exercise 3: Confidence Interval Misconceptions II

Suppose that a student is working on a statistics project using data on pulse rates collected from a random sample of 100 students from Reed college, and finds a 95% confidence interval for mean pulse rate to be 65.5 to 71.8. Discuss how each of the statements below would indicate an *improper* interpretation of this interval:



**a.** "I am 95% sure that all students will have pulse rates between 65.5 and 71.8 beats per minute."

*************************************************





************************************************* 



**b.** "I am 95% sure that the mean pulse rate for this sample of 100 students will fall between 65.5 and 71.8 beats per minute."

*************************************************





************************************************* 




**c.** "There is a 95% probability that the average pulse rate of all students at this college is in the interval from 65.5 to 71.8 beats per minute."

*************************************************





************************************************* 





**d.** "Of random samples of this size taken from students at Reed college, 95% will have mean pulse rates between 65.5 and 71.8 beats per minute."

*************************************************





************************************************* 




\newpage




## Exercise 4: Proportion of Emphathetic Rats

In a recent study, a group of rats showed compassion in a way that surprised scientists: 23 of the 30 rats in the study freed another trapped rat in their cage, even when chocolate served as a distraction and even when the rats would then have to share the chocolate with their freed companion. (Rats, it turns out, love chocolate.) Rats did not open the cage when it was empty or when there was a stuffed animal inside, only when a fellow rat was trapped. We wish to use the sample to estimate the proportion of rats that show empathy in this way.

**a.**   What population parameter do researchers wish to investigate? What is the sample statistic they can use to estimate the population parameter, and what is its value?

*********************************************************************************






*********************************************************************************



**b.** Run the following code chunk to produce the sample of 30 observations that the researchers collected and create a bootstrap distribution of 5,000 bootstrap statistics. **Then, describe the shape, center, and spread of the distribution.**

```{r fig.asp = 0.3, echo = F}
set.seed(12)
rats <- data.frame(empathy = c(rep("Yes", 23), rep("No", 7)))

rats %>% rep_sample_n(size = 30, replace = TRUE, reps = 5000) %>%
  group_by(replicate) %>%
  summarize(stat = mean(empathy == "Yes")) %>%
  ggplot(aes(x = stat)) +
  geom_histogram(col = "white", bins = 15) +
  ggtitle("Simulation-Based Bootstrap Distribution")

```


*********************************************************************************






*********************************************************************************




**c.** Based on your answer to part (b), would you have any concerns about using the **standard error bootstrap method** for confidence intervals?

*********************************************************************************






*********************************************************************************




**d.** What percentiles would you need to construct a 92% confidence interval for the parameter of interest using the percentile method? **From eyeballing the plot in part (b), give a rough estimate of those percentiles.**

*********************************************************************************
 





*********************************************************************************



**e.** Using your estimates from part (d), state and carefully interpret a 92% confidence interval for the parameter of interest.

*********************************************************************************
 





*********************************************************************************
 
 
 
**f.**	Based on the confidence interval you calculated in part (e), is there evidence that a majority of rats will show empathy?  Justify your answer.

*********************************************************************************





*********************************************************************************




## Exercise 5: Continuous Rat Empathy

Suppose rat empathy was also measured on a continuous scale from 0-100. The chunk below uses the same process as Exercise 4 (b) to create two bootstrap distributions: one where the sample statistic is the **mean empathy rating** and one where the sample statistic is the **median empathy rating**. 

```{r fig.asp = 0.3, echo = F}
set.seed(12)
rats <- data.frame(empathy = c(sample(51:100, 23, TRUE), sample(1:51, 7, TRUE)))

rats_boot <- rats %>% rep_sample_n(size = 30, replace = TRUE, reps = 5000) %>%
  group_by(replicate) %>%
  summarize(median_stat = median(empathy),
            mean_stat = mean(empathy))

gg1 <- ggplot(rats_boot, aes(x = mean_stat)) +
  geom_histogram(col = "white", bins = 15) +
  ggtitle("Sample Mean")

gg2 <- ggplot(rats_boot, aes(x = median_stat)) +
  geom_histogram(col = "white", bins = 15) +
  ggtitle("Sample Median")

ggpubr::ggarrange(gg1, gg2, nrow = 1)

```


**a.** How do these distributions differ? Should we use the **standard error bootstrap method** to generate confidence intervals for either of these?

*********************************************************************************




*********************************************************************************


**b.** Could the percentile method for confidence intervals be appropriate for either of these bootstrap distributions?

*********************************************************************************





*********************************************************************************









## Exercise 1: Hypothesis Testing Setup

For each of the situations described in parts (a) - (c), do the following:

  (i) Describe the null and alternative hypotheses (in words)
  (ii) Describe the effects of a Type I and Type II error in context
  (iii) Indicate whether it makes more sense to use a larger significance level (such as $\alpha = 0.10$) or a smaller significance level (such as $\alpha = 0.01$), and briefly explain your reasoning.


**a.** A new drug with potentially dangerous side effects is tested to see if it is better than the drug currently in use. If it is found to be effective, it will be prescribed to millions of people.

*********************************************************************************




*********************************************************************************


**b.** Two slightly different versions of a statistics exercise are given to a 10am and 12pm section of Math 141 and the proportions of students in each class who get the question correct are recorded. We wish to determine whether one question was more difficult than the other.

*********************************************************************************




*********************************************************************************


**c.** You test whether taking a vitamin supplement reduces the duration of symptoms of the common cold. There are no (known) harmful side effects of the supplement.

*********************************************************************************




*********************************************************************************



## Exercise 2: Interpreting Results and Misconceptions

After graduating, you decide to apply for positions as a statistical consultant. A friend who graduated a few years prior and already works in the field has sent you the following set of scenarios. If she is impressed by your responses, she will recommend your name to a hiring team. For each response, be sure to fully explain your reasoning.


**a.** Melatonin pills are a popular remedy for providing insomnia relief. A company developed a new formulation of melatonin pill and conducted a randomized controlled trial on a random sample of 2,000 American adults who were already experiencing insomnia and using melatonin pills. A 95% confidence interval for mean increase in sleep hours per typical night was calculated based on data from the trial: (0.09, 0.13) hours. A hypothesis test for the null hypothesis of no effect resulted in $p$-value 0.0001. **Assess whether the evidence suggests the new formulation is a substantial improvement over melatonin pills currently available on the market. Limit your response to no more than five sentences.**

*********************************************************************************




*********************************************************************************



**b.**  The company would like to announce in a press release that based on these data there is only a 0.01% chance that their new melatonin pill is not actually effective, or in other words, that there is a 99.99% chance that their new formulation on average improves the amount of sleep in American adults with insomnia. **Explain why this statement is misleading.**

*********************************************************************************




*********************************************************************************



**c.**  Suppose that a researcher in environmental science is working on an analysis using data collected at 200 locations across the United States during January of 2024 and January of 2014. The researcher proposes the following approach: for each location, conduct a two-sided hypothesis test at the $\alpha = 0.05$ significance level to compare the mean temperature in January 2024 to the mean temperature in January 2014. He plans to conclude there is evidence of temperature warming for the locations at which mean temperature in January 2024 is significantly higher than mean temperature in January 2014, and specifically present only those significant results to his principal investigator. **Identify at least one aspect of this analysis plan that is problematic; explain your reasoning as to why there is an issue.**

*********************************************************************************




*********************************************************************************


**d.**  A recent study of $n = 20$ dogs investigated the effects of classical music on dog separation anxiety (measured on a 0-10 point scale) and found inconclusive results. Researchers were unable to reject the null hypothesis of zero effect. Researchers report that classical music reduced anxiety in the sample by an effect size of -2.5, and provide a 95% confidence interval of [-6.5, 1.5] and a p-value of $p = 0.34$ for $\alpha = 0.05$. **What are 2 distinct reasons why the study failed to reject the null hypothesis, and how could researchers adjust a future study to distinguish between these 2 cases?** You can assume that the study was perfectly implemented - failure to reject the null was not due to any sampling or implementation error or accident. 

*********************************************************************************




*********************************************************************************



## Exercise 3: P-Value Pitfalls

The pitfalls of $p$-values have been discussed for decades, and in 2016 the *ASA Statement on p-Values and Statistical Significance* was developed primarily to discuss what not to do with $p$-values.


**a.**  To better understand the debate around the use of $p$-values, read the article "A Significant Problem" (written by Lydia Denworth) in the file `scientific_american_p_values.pdf`, available here with your assignment. (*No response required for part a.*)



**b.**  Describe one reason why it is not advisable to label all results with p-values $p < 0.05$ as "significant" and all results with p-values $p \geq 0.05$ as not significant.

*********************************************************************************




*********************************************************************************




**c.**  Take a look at the section "Statistical Significance" on pages 2-4, which is intended to explain some concepts underpinning statistical significance in language accessible to those who have not taken a statistics course. **Describe in 2-3 sentences something you thought was explained/illustrated particularly well in this section.**

*********************************************************************************




*********************************************************************************



**d.** Try doing some $p$-hacking of your own with ["Hack Your Way To Scientific Glory"](https://stats.andrewheiss.com/hack-your-way/) to reach a "publishable" result. **How did you get to the statistically "significant" result?**

*********************************************************************************




*********************************************************************************


**e.**  One proposed reform involves requiring study authors to publicly pre-register their statistical analysis plans prior to any data collection and to report when they deviate from them. **Explain in 2-4 sentences why this approach can help guard against $p$-hacking.**

*********************************************************************************





*********************************************************************************


