---
title: "Lab 5: Multiple Linear Regression"
subtitle: "STAT 141, Week 5"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

In this lab, you will

1. Review exploratory data analysis (EDA) techniques in R with multiple predictors
2. Assess linear regression models with multiple predictors
3. Interpret the estimated coefficients of multiple linear regression models
4. Compare and assess linear regression models


### Add any collaborators here!

I collaborated with...

**Remember that you are not allowed to use any internet resources or generative AI models (like ChatGPT or Claude) on your assignment. While there are many ways to achieve tasks in R, you may only use packages and functions that have been covered in this course. If you are unsure of how to approach an exercise or of what function to use, help is available to you! Please visit office hours, collaborate with your peers, or post in the Slack channel.**


Run the chunk below to load the necessary packages.

```{r}
library(tidyverse)
library(moderndive)
```

# Exercises

## Exercise 1: Exploratory Data Analysis for Regression

We will study data on teacher salaries for this lab assignment. Our data is a sample of 100 post-secondary (i.e., college) teachers is collected from the 2010 American Community Survey Public Use Microdata Sample, as found in the `Lock5Data` R package. We are interested in exploring differences in salary based on gender (i.e., gender-based salary discrimination) among college teachers.

For this exercise, we will focus on the role of exploratory data analysis for linear regression models.

Load the dataset using the following code:

```{r}
library(Lock5Data)
data("SalaryGender")
SalaryGender <- SalaryGender %>%
  select(Gender, Age, Salary) %>%
  mutate(Gender = ifelse(Gender == 0, "Female", "Male"))
```

The data includes three variables:

1. **Gender**: Coded as a binary "Female" or "Male". It is not clear whether this truly reflects the gender identity of participants, or something different like biological sex. (We might assume that no participants with other gender identities were included, which limits our ability to generalize to the population at large.)
2. **Age**: Age of participant, in years.
3. **Salary**: Annual salary in thousands of US dollars.

The distributions of these three variables are visualized below individually.

```{r}
g1 <- ggplot(data = SalaryGender, aes(x = Salary)) + geom_histogram(bins = 20) +
  theme_bw() + labs(title = "Histogram of Salary", x = "Salary (thousands of dollars)")
g2 <- ggplot(data = SalaryGender, aes(x = Age)) + geom_histogram(bins = 20) +
  theme_bw() + labs(title = "Histogram of Age", x = "Age (years)")
g3 <- ggplot(data = SalaryGender, aes(x = Gender)) + geom_bar() +
  theme_bw() + labs(title = "Barchart of Gender")
gridExtra::grid.arrange(g1, g2, g3, nrow = 2)
```



**a.** The chunk below investigates (1) the bivariate relationship between `Salary` and `Gender,`  (2) the bivariate relationship between `Salary` and `Age`, and (3) the bivariate relationship between `Age` and `Gender`.  For **each of the three graphs**, explain in 1-2 sentences **what trends you observe, and how they are relevant to building our linear model.** Keep in mind that our overall goal is to understand the relationship between salary and gender. [Text only for this question]

```{r}
ggplot(data = SalaryGender, aes(x = Gender, y = Salary)) + 
  geom_boxplot() + theme_bw() + 
  labs(title = "Side-by-Side Boxplots of Salary by Gender")

ggplot(SalaryGender, aes(x = Age, y = Salary)) + 
  geom_point() + geom_smooth(method = "lm", se = F) + theme_bw() +
  labs(title = "Scatterplot of Salary by Age")

ggplot(data = SalaryGender, aes(x = Gender, y = Age)) + 
  geom_boxplot() + theme_bw() + 
  labs(title = "Side-by-Side Boxplots of Age by Gender")

```

*******************************************************************************

(1)


(2)


(3)


*******************************************************************************



**b.** In a new chunk below, create a scatterplot of salary vs age, colored by gender. Add gender-specific trend lines by adding `+ geom_parallel_slopes(se = FALSE)`. **Comment on any trends observed.** [Code and text for this question]

*******************************************************************************



*******************************************************************************



## Exercise 2: Salary Regression Model

In this exercise, you will use a regression model to investigate whether age explains the gap in salary between males and females in post-secondary education. (Note: Even *if* age explained the gap, **there is still a gap!** This would be an indication of structural differences for men and women in the US.)

**a.** In the chunk below, create a *multiple linear regression model* that predicts `Salary` using both `Gender` and `Age`, with fixed slopes. Display the model's regression summary table. Using the summary table, **(1)** write the equation for the fitted model, and **(2)** interpret each coefficient, explaining how their values validate your exploratory findings from Exercise 1. [Code and text needed for this question]

*******************************************************************************

```{r}
# Create your model and get its regression table here

```


(1)


(2)


*******************************************************************************



**b.** Using your regression equation from above, calculate, in `R`, the model's prediction for the salary of a female teacher who is 30 years old. **Based on the exploratory visualizations in Exercise 1, is this an example of extrapolation?** [Code and text for this question]

*******************************************************************************



*******************************************************************************



## Exercise 3: Considering Other Models

The model created for the previous exercise is just one example of a model that we could fit using the data set. The chunk below creates two more models: `m1`, which predicts `Salary` using only `Gender`, and `m2`, which predicts `Salary` using only `Age`.

```{r}
m1 <- lm(Salary ~ Gender, data = SalaryGender)
m2 <- lm(Salary ~ Age, data = SalaryGender)
m3 <- lm(Salary ~ Age + Gender, data = SalaryGender)
```


**a.** Which of these three models helps us best answer our original research question? **Consider both $R^2$ and the types of relationships each model captures in your argument.** Recall that you can find the $R^2$ of a model by using `get_regression_summaries()`. [Code and text for this question]

*******************************************************************************



*******************************************************************************



**b.** Building on our model from 2(b) (aka model `m3`), create one more multiple linear regression model that predicts `Salary` using `Age` and `Gender`, this time allowing for *varying slopes*. How did the coefficient estimates change? [Code and text for this question]

*******************************************************************************




*******************************************************************************



**c.** Create a `ggplot` graph that visualizes the varying slopes model that you just build in part 3(b). (*Hint: compare examples of `geom_parallel_slopes()` and `geom_smooth()` in the slides.*)

*******************************************************************************




*******************************************************************************



**d. (BONUS!)** To further investigate each of these 4 models, explore their diagnostic plots. Based on the data context and the graphs produced in the chunk below, does any model meet all LINE assumptions? [Text only for this question]

*******************************************************************************




*******************************************************************************


