---
title: "Lab 5: Multiple Linear Regression"
subtitle: "STAT 141, Week 5"
author: "Your name here"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

In this lab, you will

1. Conduct exploratory data analysis (EDA) techniques in R with multiple predictors
2. Build linear regression models with multiple predictors
3. Interpret the estimated coefficients of multiple linear regression models
4. Compare and assess multiple linear regression models


### Add any collaborators here!

I collaborated with...

**Remember that you are not allowed to use any internet resources or generative AI models (like ChatGPT or Claude) on your assignment. While there are many ways to achieve tasks in R, you may only use packages and functions that have been covered in this course. If you are unsure of how to approach an exercise or of what function to use, help is available to you! Please visit office hours, collaborate with your peers, or post in the Slack channel.**

```{r}
#load the necessary packages
library(tidyverse)
library(pdxTrees)
library(moderndive)
```

# Exercises

## Exercise 1: Exploratory Data Analysis for Regression

We will study data on teacher salaries for the rest of this homework assignment. Our data is a sample of 100 post-secondary (i.e., college) teachers is collected from the 2010 American Community Survey Public Use Microdata Sample, as found in the `Lock5Data` R package. We are interested in exploring differences in salary based on gender (i.e., gender-based salary discrimination) among college teachers.

For this exercise, we will focus on the role of exploratory data analysis for linear regression models.

Load the dataset using the following code:

```{r}
library(Lock5Data)
data("SalaryGender")
SalaryGender <- SalaryGender %>%
  select(Gender, Age, Salary) %>%
  mutate(Gender = ifelse(Gender == 0, "Female", "Male"))
```

The data includes three variables:

1. Gender: Coded as a binary "Female" or "Male". It is not clear whether this truly reflects the gender identity of participants, or something different like biological sex. (We might assume that no participants with other gender identities were included, which limits our ability to generalize to the population at large.)
2. Age: Age of participant, in years.
3. Salary: Annual salary in thousands of US dollars.


**a.** As stated above, we are interesting in exploring salary differences based on gender. It's important to start by thinking critically about the type of conclusions we can draw from this analysis, especially since we're analyzing data about a sensitive (and very important) topic. **What type of study is this data from, and what types of conclusions can be drawn as a result?**

*******************************************************************************



*******************************************************************************

**b.** The chunk below creates exploratory visualizations of each variable: `Salary`, `Age`, and `Gender`. **Why is it important to perform this step before building a linear regression model? Based on the visualizations, give an insight about each variable.**

```{r}
g1 <- ggplot(data = SalaryGender, aes(x = Salary)) + geom_histogram(bins = 20) +
  theme_bw() + labs(title = "Histogram of Salary", x = "Salary (thousands of dollars)")
g2 <- ggplot(data = SalaryGender, aes(x = Age)) + geom_histogram(bins = 20) +
  theme_bw() + labs(title = "Histogram of Age", x = "Age (years)")
g3 <- ggplot(data = SalaryGender, aes(x = Gender)) + geom_bar() +
  theme_bw() + labs(title = "Barchart of Gender")
gridExtra::grid.arrange(g1, g2, g3, nrow = 2)
```

*******************************************************************************



*******************************************************************************

**c.** The chunk below investigates the bivariate relationship between `Salary` and `Gender.` **What trends do you observe, and how are they relevant to building our linear model?**

```{r}
ggplot(data = SalaryGender, aes(x = Gender, y = Salary)) + geom_boxplot() +
  theme_bw() + labs(title = "Side-by-Side Boxplots of Salary by Gender")
```

*******************************************************************************



*******************************************************************************

**d.** Assess the scatterplot produced by the chunk below. **Describe what the graph is showing.**

```{r}
ggplot(SalaryGender, aes(x = Age, y = Salary)) + geom_point() +
  geom_smooth(method = "lm", se = F) + theme_bw()
```

*******************************************************************************



*******************************************************************************

**e.** Calculate the correlation coefficient between `Salary` and `Age`, and **comment on the strength of the relationship**.

*******************************************************************************



*******************************************************************************

**f.** Assess the boxplots produced by the chunk below. **Describe what the graph is showing. Explain how insights from this graph and the graph from part (d) are relevant to building our linear model.** Keep in mind that our overall goal is to understand the relationship between salary and gender

```{r}
ggplot(SalaryGender, aes(x = Gender, y = Age)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = "Side-by-Side Boxplots of Age by Gender")
```

*******************************************************************************



*******************************************************************************

**g.** In a new chunk below, create a scatterplot of salary vs age, colored by gender. Add gender-specific trend lines by adding `+ geom_parallel_slopes(se = FALSE)`. **Comment on any trends observed.**

*******************************************************************************



*******************************************************************************

## Exercise 2: Salary Regression Model

In this exercise, you will use a regression model to investigate whether age explains the gap in salary between males and females in post-secondary education. (Note: Even *if* age explained the gap, **there is still a gap!** This is an indication of structural differences for men and women in the US.)

**a.** The chunk below creates a linear model using the `SalaryGender` data frame. **Using the summary table, write the equation for the fitted model. Interpret each coefficient, and explain how their values validate your exploratory findings from Exercise 1.**

```{r}
lm_salary_gender <- lm(Salary ~ Age + Gender, data = SalaryGender)
get_regression_table(lm_salary_gender)
```

*******************************************************************************



*******************************************************************************

**b.** Using your regression equation from part (a) of this exercise, calculate the model's prediction for the salary of a female teacher who is 30 years old. **Based on the exploratory visualizations in Exercise 1, is this an example of extrapolation?**

*******************************************************************************



*******************************************************************************

## Exercise 3: Assess Salary Regression Models

**a.** The model created for the previous exercise is just one example of a model that we could fit using the data set. **Briefly explain how each of the four models below is different from each other.**

```{r}

m1 <- lm(Salary ~ Gender, data = SalaryGender)
m2 <- lm(Salary ~ Age, data = SalaryGender)
m3 <- lm(Salary ~ Age + Gender, data = SalaryGender)

```

*******************************************************************************



*******************************************************************************

**b.** Which of the three models helps us best answer our original research question? **Consider both $R^2$ and the types of relationships each model captures in your argument.**

*******************************************************************************



*******************************************************************************

**c.** Investigating when linear regression model assumptions hold is important to consider when selecting the best model(s) from a collection of possible models. Run the chunk below, and assess the set of graphs shown. **Which LINE assumption(s) do these graphs relate to, and which model(s) satisfy it/them?**

```{r}
# Use get_regression_points to grab preditions (aka "fitted values"), residuals
# for each model.
m1_pts <- get_regression_points(m1)
m2_pts <- get_regression_points(m2)
m3_pts <- get_regression_points(m3)


# These next two commands are not usually needed, but here they help us compare
# the three different models directly within the same data frame (and therefore
# the same ggplot).
res <- bind_rows(m1_pts %>% mutate(model = "Salary ~ Gender") %>%
                   select(Salary, Salary_hat, residual, model),
                 m2_pts %>% mutate(model = "Salary ~ Age") %>%
                   select(Salary, Salary_hat, residual, model),
                 m3_pts %>% mutate(model = "Salary ~ Age + Gender") %>%
                   select(Salary, Salary_hat, residual, model))
res$model <- factor(res$model, levels = c("Salary ~ Gender",
                                          "Salary ~ Age",
                                          "Salary ~ Age + Gender"))

ggplot(res, aes(x = Salary_hat, y = residual)) +
  geom_hline(aes(yintercept = 0), lty = 2, color = "gray") +
  geom_point() +
  facet_wrap(~model) +  # This line shows us one plot per model - usually we don't include it
  labs(title = "Residuals vs Fitted Values") +
  theme_classic()

```

*******************************************************************************



*******************************************************************************

**d.** Run the chunk below, and assess the next set of diagnostic graphs shown. **Which LINE assumption(s) do these graphs relate to, and which model(s) satisfy it/them?**

```{r}

ggplot(res, aes(x = residual)) +
  geom_histogram(bins = 20, color = "white") +
  facet_wrap(~model) +  # This line shows us one plot per model - usually we don't include it
  labs(title = "Residual Histograms") +
  theme_classic()

```

*******************************************************************************



*******************************************************************************

**e.** Would you feel comfortable using any of these models to argue that there is evidence of gender-based salary discrimination among college teachers in 2010?

*******************************************************************************



*******************************************************************************
