---
title: "Multiple Linear Regression: Inference and Model Selection"
format: pdf
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = 'center')
library(tidyverse)
library(moderndive)
library(gridExtra)
set.seed(1)
data <- nycflights13::flights %>%
  filter(dest == "PDX", arr_delay < 750)
mod <- lm(arr_delay ~ air_time + dep_delay * carrier, data)
data <- mod %>% get_regression_points()
data$arr_delay <- data$arr_delay_hat + rnorm(1341, 0, sd = 25)
data <- data %>% select(arr_delay, carrier, dep_delay, air_time)
data$carrier <- factor(data$carrier, levels = c("B6", "DL", "UA"),
                       labels = c("JetBlue", "Delta", "United"))
rm(mod)
```



In this worksheet, we'll practice concepts related to inference for regression coefficients, F-Tests, and model selection. Work through the worksheet carefully in small groups.



## Data: Flights to PDX Airport
We will analyze data on flights arriving to the PDX airport in 2013. We have data on $n=1341$ randomly sampled flights, and for each flight, we have 4 variables:


+ Response: Arrival Delay in minutes (`arr_delay`). Can be negative if flight arrived early.
+ Explanatory 1: Airline Carrier (`carrier`).
+ Explanatory 2: Departure Delay in minutes (`dep_delay`). Can be negative if the flight departed early.
+ Explanatory 3: Air Time in minutes (`air_time`). 

```{r, out.width="100%", fig.asp=0.5}
grid.arrange(
  ggplot(data, aes(x = arr_delay)) + geom_histogram() + labs(title = "Fig. 1") +
    theme_bw(),
  ggplot(data, aes(x = carrier, y = arr_delay)) + geom_boxplot() + 
    labs(title = "Fig. 2") + theme_bw(),
  ggplot(data, aes(x = dep_delay, y = arr_delay)) + geom_point(alpha = 0.3) +
    labs(title = "Fig. 3") + theme_bw(),
  ggplot(data, aes(x = air_time, y = arr_delay)) + geom_point(alpha = 0.3) +
    labs(title = "Fig. 4") + theme_bw(),
  nrow = 2
)
```


### Question 1:


Which variables seem to be important predictors of arrival delay? Consider the plots above and context.

*******************************************************************************



*******************************************************************************



\newpage




### Question 2:


Let's test if arrival delay varies by airline carrier using ANOVA. 

**a.** Specify the relevant parameters and the set of hypotheses.

*******************************************************************************


  
*******************************************************************************
  
  






**b.** State the p-value calculated by R (below) and interpret the result in context. What have you learned about these airline carriers from the ANOVA F-test and Figure 2?

```{r echo=T}
mod_carrier <- lm(arr_delay ~ carrier, data)
anova(mod_carrier)
```

*******************************************************************************



*******************************************************************************



\newpage




### Question 3:


Suppose we decide to model arrival delay using air time, departure delay, and carrier. We are considering additionally including an interaction term between departure delay and carrier. We will use a Nested F-Test to determine if we should use the "small" or "large" model, as defined below.

  + Small Model: Predict Arrival Delay with Air Time, Departure Delay, and Carrier
  + Large Model: Predict Arrival Delay with Air Time, Departure Delay, Carrier, and the interaction between Departure Delay and Carrier. 

**a.** How many coefficients are in each of the small and large models? What variables do they represent?

*******************************************************************************



*******************************************************************************





**b.** Write down the set of hypotheses for a Nested F-Test to compare the small and large model above.

*******************************************************************************


  
*******************************************************************************
  
  
  


**c.** State the p-value calculated by R (below) and interpret the results of the Nested F-Test in context.

```{r echo=T}
mod1 <- lm(arr_delay ~ air_time + dep_delay + carrier, data)
mod2 <- lm(arr_delay ~ air_time + dep_delay * carrier, data)
anova(mod1, mod2)
```

*******************************************************************************



*******************************************************************************


\newpage



### Question 4:


Suppose your colleague proposes using a Nested F-Test to determine compare the two models defined below.

  + Model 1: Predict Arrival Delay with Departure Delay, log(Departure Delay), and Carrier
  + Model 2: Predict Arrival Delay with Air Time, Carrier, and the interaction between Air Time and Carrier. 

Explain why a nested F-test \underline{cannot} be conducted to compare these two models.

*******************************************************************************




*******************************************************************************




\newpage




### Question 5:
Below are two plots to help assess assumptions of the large model, as defined before in Q3. Assess the conditions for conducting theory-based inference with linear regression.

```{r out.width="100%",fig.asp=0.4}
grid.arrange(
  mod2 %>% get_regression_points() %>%
    ggplot(aes(x = arr_delay_hat, y = residual)) + geom_point(alpha = 0.3) +
    geom_smooth(method = "lm", se = F) + theme_bw(),
  mod2 %>% get_regression_points() %>%
    ggplot(aes(x = residual)) + geom_histogram(bins = 30) + theme_bw(),
  nrow = 1
)
```

*******************************************************************************





*******************************************************************************


\newpage



### Question 6:

Examine the regression table below.

```{r echo=T}
get_regression_table(mod2, conf.level = 0.8)
```

**a.** Write down the estimated model equation implied by the regression table.

*******************************************************************************





*******************************************************************************





**b.** Interpret, in context, the coefficients for the (i) intercept, (ii) slope term on air time, (iii) the slope term on departure delay. Be particularly careful for when interpreting (iii) !

*******************************************************************************





*******************************************************************************




**c.** Write down the equation for the specific case when the airline carrier is Delta.

*******************************************************************************





*******************************************************************************




**d.** What is the confidence level for the confidence intervals shown? 

*******************************************************************************





*******************************************************************************


