---
title: "Case Study: Home Prices in King County, WA"
format: pdf
---

\vspace{-2.5cm}

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = 'center')
library(tidyverse)
library(moderndive)
library(gridExtra)
house <- read_csv("data/house.csv")
```

## What's the relationship between a home's square footage and number of bedrooms on the home's sale price?

We'll consider \textbf{four variables}: Sale price (**response**); and square footage, above-ground square footage, and number of bedrooms (potential explanatory variables). Check out the exploratory plots below:


```{r outwidth="100%",fig.asp=0.65}

# Distribution of price
gg1 <- ggplot(house, aes(x = price)) + 
  geom_histogram(binwidth = 25, color = "white") + 
    theme_bw(base_size = 11) +
    labs(x = "Price (thousands of dollars)")

# Distribution of bedrooms
gg2 <- ggplot(house, aes(x = bedrooms)) + 
  geom_histogram(binwidth = 1, color = "white") + 
    theme_bw(base_size = 11) +
    labs(x = "Number of Bedrooms") + 
  scale_x_continuous(breaks = 1:10)

# Distribution of sqft_living
gg3 <- ggplot(house, aes(x = sqft_living)) + 
  geom_histogram(binwidth = 100, color = "white") + 
    theme_bw(base_size = 11) +
    labs(x = "Total Square Footage")

# Distribution of sqft_above
gg4 <- ggplot(house, aes(x = sqft_above)) + 
  geom_histogram(binwidth = 100, color = "white") + 
    theme_bw(base_size = 11) +
    labs(x = "Above-Ground Square Footage")

grid.arrange(gg1, gg2, gg3, gg4, nrow = 2)  # Arranges our 4 graphs nicely
```


- **Q**: How would you describe the distribution of `price`?

- **Q**: What do you expect the relationship between `price` and `bedrooms` to be like?

\newpage


## Exploration, Multicollinearity

It's important to look at relationships **between explanatory variables.**

```{r outwidth="100%",fig.asp=0.45}

grid.arrange(  # grid.arrange places all three of these graphs in a row for us
  
  # Plots sqft_above vs sqft_living 
  ggplot(house, aes(x = sqft_living, y = sqft_above)) + 
    geom_point(alpha = 0.3) + theme_bw(base_size = 10) +
    labs(x = "Sq. Footage", y = "Ab.-Ground Sq. Ft."),
  
  # Plots bedrooms vs sqft_living 
  ggplot(house, aes(x = sqft_living, y = bedrooms)) + 
    geom_point(alpha = 0.3) + theme_bw(base_size = 10) +
    labs(x = "Sq. Footage", y = "Bedrooms"),
  
  # Plots bedrooms vs sqft_above 
   ggplot(house, aes(x = sqft_above, y = bedrooms)) + 
    geom_point(alpha = 0.3) + theme_bw(base_size = 10) +
    labs(x = "Ab.-Ground Sq. Ft.", y = "Bedrooms"),
  nrow = 1)

```

- **Multicollinearity**: when explanatory variables are **highly** correlated with one another. Multicollinearity often results in coefficients that are distorted in erroneous ways! We want to avoid this (*some* correlation is okay!)

- **Q**: Which pair of explanatory variables suffers *most* from multicollinearity?

\newpage

We also should check relationships **between potential explanatory variables and the outcome** to assess **linearity**.

```{r outwidth="100%",fig.asp=0.75}

### Can use ggpairs() from the GGally library to visualize ALL pairwise and
### individual distributions.
library(GGally)
house %>% select(sqft_living, sqft_above, bedrooms, price) %>%
  ggpairs()
```

- **Q**: Linear relationships with price?

\newpage


## Model Estimation

```{r}
mod <- lm(price ~ sqft_living + bedrooms, data = house)  # Builds model
get_regression_table(mod)  # Gets regression summary table
```

- **Q**: What is the equation, including coefficient values, for the estimated model?


- **Q**: How should interpret the **intercept** coefficient in our model?


- **Q**: The coefficient for **Bedrooms**? Does this match your expectation?


\newpage



## LINE Assumptions

```{r}
res <- get_regression_points(mod)  # Gets regression residuals, fitted values

res_plot <- ggplot(res, aes(x = price_hat, y = residual)) +
  geom_point() + geom_hline(aes(yintercept = 0), lty = 2, color = "gray50")

res_hist <- ggplot(res, aes(x = residual)) + geom_histogram(color = "white")

qq_gg <- ggplot(res, aes(sample = residual)) + geom_qq() + geom_qq_line()

grid.arrange(res_plot, res_hist, qq_gg, nrow = 2)  # Display diagnostics together

```
- **Q**: Are the LINE assumptions met? 


\newpage


## Adjusted $R^2$

What's the adjusted $R^2$ of our model?

```{r echo = TRUE}
# get_regression_summaries gives summary stats about the model as a whole,
# including R^2 and adjusted R^2. The brackets grab just these two results.
get_regression_summaries(mod)[, c("r_squared", "adj_r_squared")]
```

Not super close to 1. Compare to a simpler model:



```{r echo = TRUE}
small_mod <- lm(price ~ bedrooms, data = house)
get_regression_summaries(small_mod)[, c("r_squared", "adj_r_squared")]
```



Compare to a much bigger model with additional variables:

```{r echo = TRUE}
big_mod <- lm(price ~ bedrooms + bathrooms + sqft_living + sqft_above + 
                sqft_lot + view + condition + yr_built, data = house)
get_regression_summaries(big_mod)[, c("r_squared", "adj_r_squared")]
```


\newpage


## Chunk error example

```{r}
# lm(price ~ Bedrooms, data = house)
```

